Automatic Identification and Annotation of Tasks in Structured Programs

### 2018-04-02 docs/ACM/paper.tex

\paragraph{Tasks: syntax and scope:}
Our goal is to identify {\em tasks} in {\em structured} programs.
A structured program is a program that can be partitioned in {\em hammock
regions}, a concept introduced by Ferrante {\em et al.}~\cite{Ferrante87} in
the middle 80's.
For completeness, we re-state it in Definition~\ref{def:hammock}.
Definition~\ref{def:task} formalizes the notion of task.

\begin{definition}[Hammock Region~\cite{Ferrante87}]
\label{def:hammock}
A Control Flow Graph $G$ is a directed graph with a starting node $s$, and an
exit node $x$.
A hammock region $G$ is a subgraph of $G$ with a node $h$ that
{\em dominates}\footnote{A node $n_1 \in G$ dominates another node
$n_2 \in G$ if every path from  $S$ to $n_2$ goes across $n_1$.
Inversely, $n_1$ post-dominates $n_2$ if every path from $n_2$ to $x$ must
go across $n_1$.} the other nodes in $G'$.
Additionally, there exists a node $w \in G$, $w \notin G'$, such that
$w$ {\em post-dominates} every node in $G'$.
In this definition, $h$ is the {\em entry point}, and $w$ is the {\em exit point}
of $G'$.
\end{definition}

@@

\paragraph{The annotation zoo.}
Our equipment of choice to reveal task parallelism in programs is OpenMP 4.0.
%This annotation system lets us indicate to compiler
%and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations is publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by one thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default([shared/private])}: indicates that variables are either
shared or replicated among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause defines the conditions
under which a task is allowed to create threads.
We use it to enforce the cost model.
\end{compactitem}

@@

% How we do it.
The program symbols mentioned in Definition~\ref{def:sra} are names that cannot be
reconstruct as functions of other names.
Examples include global variables, function arguments, and values returned by
external functions.
Symbolic range analysis is not a contribution of this paper.
There are several different implementations publicly available, and we have
adopted the one in the \dawn{} compiler~\cite{Mendonca17}, which, itself,
reuses work from Blume {\em et al.}~\cite{Blume94}.
The only extensions that we have added into \dawn's implementation was
the ability to handle \textsf{structs}.
Therefore, we shall not provide further details about part of our work.
To understand the rest of this paper, it suffices to know that this implementation
is sufficiently solid to handle the entire C99 language, always terminates, and
runs in time linear on the size of the program's dependence graph.
Thus, in the worst case, it is quadratic on the number of program variables.
If a program contains an array access $\textsf{V}[E]$, and we have that
$R(E) = [l, u]$, then the memory region covered by this access is, at least,
$[\mathtt{\&}\textsf{V} + l, \mathtt{\&}\textsf{V} + u]$.
The example below clarifies this observation.

@@

\begin{definition}[Windmill -- \cite{Rideau08}]
\label{def:windmill}
A windmill is a graph $G_w = G_c \cup G_{v1} \cup \ldots \cup G_{vn}$ formed by
a strongly connected component $G_c$, its center, plus $n$ components (not
necessarily strong) $G_{vi}$, the vanes, such that:
\begin{compactenum}
\item for any topological ordering of $G_w$, and nodes $n_c \in G_c$,
$n_v \in G_{vi}$, we have $n_c$ ahead of $n_v$;
\item for any $i$ and $j$, $1 \leq i < j \leq n$, $G_{vi}$ and $G_{vj}$ do not
share vertices (thus, sharing edges is also impossible).
\end{compactenum}
\end{definition}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\columnwidth]{images/expand_alg}
\caption{Task discovery via expansion of hammock regions.
\textsf{COST} is the overhead of creating and scheduling threads, as determined
by a cost model.}
\label{fig:expand_alg}
\end{center}
\end{figure}

The boolean \textsf{CanExpand} in Figure~\ref{fig:expand_alg} is determined by
the routine \textsf{checkExpansion}.
This function remains true for as long as the following conditions apply with
respect to its input \textsf{VANE}:
(i) \textsf{VANE} is a hammock region with the structural properties of a
windmill's vane, i.e., it is contained within a windmill $W$;
(ii) we can analyze all the memory regions within \textsf{VANE} using the
symbolic range analysis of Section~\ref{sub:sra};
(iii) \textsf{VANE} does not depend on another vane inside the same windmill $W$.
In the rest of this section, we shall use Figure~\ref{fig:ex_expansion} to
explain these conditions.

@@

The program in Figure~\ref{fig:ex_expansion} (a), which is a slight variation of
the code earlier seen in Figure~\ref{fig:ex_regions}, contains two windmills.
The first consists in the \textsf{for} loop at line 1.
The second, nested in the first, is the \textsf{for} loop at line 2.
The vanes that sprout from these windmills, are highlighted in the program
dependence graph seen in Figure~\ref{fig:ex_expansion} (b).
The figure shows that we have three task candidates, each one formed by a
different vane.
In order to find the actual tasks, we apply function \textsf{expand}
(Fig.~\ref{fig:expand_alg}) onto the innermost vanes: \textsf{Vane-2} and
\textsf{Vane-3}.
The latter cannot be further expanded, because it depends on the former.
On the other hand, \textsf{Vane-2} can be expanded to include also
\textsf{Vane-3}, as this expansion does not create new dependences between vanes
in the same windmill.

\subsection{Limiting the Creation of Recursive Tasks}
\label{sub:rec}

Although the OpenMP runtime guarantees correctness, its scheduling policy
is heavily jeopardized by the excessive queueing up of tasks. Specifically, algorithms
such as \emph{SparseLU} and \emph{FFT} can suffer bloat because of
too many tasks in the Task Graph\cite{Huynh17}.

%\pedro{In the Kenjiro's paper: "This scheduling policy suffers in two situations: 
%(1) the parent is a long for loop creating a large number of children, 
%all these children cannot start until the parent finishes the task-generating loop (SparseLU); 
%(2) binary recursions that, rightly, spawn only the first recursive call as a task (FFT). 
%It imposes an upper bound for the ready task queue of a worker; a worker 
%is discouraged to create too many tasks ready on the queue"}.

To solve this problem, we conservatively limit the number of tasks
that can be created tied to the same Task Region. We do that by two means:
(1) limiting the creation of recursive tasks by depth; (2) limiting the creation of 
tasks in the whole program.

Consider the FIB

\pedro{Show FFT example}

\pedro{Explain that the setting is conservative and thus the parameter of control
(max recursion depth and total number of tasks) can be easily switched by the
user, or even completely disabled.}

\subsection{Separating Private from Shared Variables}
\label{sub:variance}

@@

>>>>

\paragraph{Tasks: syntax and scope:}
Our goal is to identify {\em tasks} in {\em structured} programs.
In the context of this work, a structured program is a program that can be
partitioned in {\em hammock regions}, a concept introduced by Ferrante
{\em et al.}~\cite{Ferrante87} in the middle 80's.
For completeness, we re-state it in Definition~\ref{def:hammock}.
Definition~\ref{def:task} formalizes the notion of task.

\begin{definition}[Hammock Region~\cite{Ferrante87}]
\label{def:hammock}
A Control Flow Graph $G$ is a directed graph with a starting node $s$, and an
exit node $x$.
A hammock region $G'$ is a subgraph of $G$ with a node $h$ that
{\em dominates}\footnote{A node $n_1 \in G$ dominates another node
$n_2 \in G$ if every path from  $S$ to $n_2$ goes across $n_1$.
Inversely, $n_1$ post-dominates $n_2$ if every path from $n_2$ to $x$ must
go across $n_1$.} the other nodes in $G'$.
Additionally, there exists a node $w \in G$, $w \notin G'$, such that
$w$ {\em post-dominates} every node in $G'$.
In this definition, $h$ is the {\em entry point}, and $w$ is the {\em exit point}
of $G'$.
\end{definition}

@@

\paragraph{The annotation zoo.}
Our equipment of choice to exploit task parallelism in programs is OpenMP 4.0.
%This annotation system lets us indicate to compiler
%and runtime environment when and how tasks must be created.
We describe below which annotations we are using, and explain, informally, their
semantics.
Full overviews of the syntax and semantics of these annotations is publicly
available\footnote{For a quick overview, we refer the reader to the leaflet
``Summary of OpenMP 4.0 C/C++ Syntax", which is made available by the OpenMP
group.}; hence, we shall not dive into their details.
%
\begin{compactitem}
\item \textsf{parallel} (\textit{clauses}): forms a team of threads
which will execute the marked program region in parallel.
\item \textsf{single} (\textit{clauses}): specifies that a
program region must be executed by one thread in the team.
\item \textsf{task} (\textit{clauses}): creates a new task.
\item \textsf{taskwait}: defines a synchronization point where
threads must wait for the completion of child tasks.
\end{compactitem}
%
Some pragmas are associated with lists of clauses.
There are several of these clauses available in OpenMP 4.0; however, we only use
the following:
\begin{compactitem}
\item \textsf{default([shared/private])}: indicates that variables are either
shared or replicated among tasks.
\item \textsf{firstprivate($v$)}: indicates that $v$ must be replicated among
tasks, and initialized with its value at the point where the annotation is
executed.
\item \textsf{untied}: if a task has this modifier, then its associated code can
be executed by more than one thread; e.g., threads might alternate execution due
to preemption and load balancing.
\item \textsf{depend}($\mathit{in}/\mathit{out}/\mathit{inout}$):
determine if data is read ($\mathit{in}$), written ($\mathit{out}$) or both
within a task region; thus, effectively setting dependences among tasks.
\item \textsf{if}({\em condition}): this clause defines the conditions
under which a task is allowed to create threads.
We use it to enforce the cost model.
\end{compactitem}

@@

% How we do it.
The program symbols mentioned in Definition~\ref{def:sra} are names that cannot be
reconstruct as functions of other names.
Examples include global variables, function arguments, and values returned by
external functions.
Symbolic range analysis is not a contribution of this paper.
There are several different implementations publicly available, and we have
adopted the one in the \dawn{} compiler~\cite{Mendonca17}, which, itself,
reuses work from Blume {\em et al.}~\cite{Blume94}.
The only extensions that we have added into \dawn's implementation was
the ability to handle C-like \textsf{structs}.
Therefore, we shall not provide further details about this part of our work.
To understand the rest of this paper, it suffices to know that this implementation
is sufficiently solid to handle the entire C99 language, always terminates, and
runs in time linear on the size of the program's dependence graph.
Thus, in the worst case, it is quadratic on the number of program variables.
If a program contains an array access $\textsf{V}[E]$, and we have that
$R(E) = [l, u]$, then the memory region covered by this access is, at least,
$[\mathtt{\&}\textsf{V} + l, \mathtt{\&}\textsf{V} + u]$.
The example below clarifies this observation.

@@

\begin{definition}[Windmill -- \cite{Rideau08}]
\label{def:windmill}
A windmill is a graph $G_w = G_c \cup G_{v1} \cup \ldots \cup G_{vn}$ formed by
a strongly connected component $G_c$ (its center), plus $n$ components, not
necessarily strong, $G_{vi}$, the vanes, such that:
\begin{compactenum}
\item for any topological ordering of $G_w$, and nodes $n_c \in G_c$,
$n_v \in G_{vi}$, we have $n_c$ ahead of $n_v$;
\item for any $i$ and $j$, $1 \leq i < j \leq n$, $G_{vi}$ and $G_{vj}$ do not
share vertices (thus, sharing edges is also impossible).
\end{compactenum}
\end{definition}

@@

\begin{figure}[h]
\begin{center}
\includegraphics[width=1\columnwidth]{images/expand_alg}
\caption{Task discovery via expansion of hammock regions.
\textsf{COST} is the overhead of creating and scheduling threads.}
\label{fig:expand_alg}
\end{center}
\end{figure}

The routine \textsf{checkExpansion} sets the boolean \textsf{CanExpand} in
Figure~\ref{fig:expand_alg}.
This function is true as long as the following conditions apply:
(i) \textsf{VANE} has the structural properties of a
windmill's vane, i.e., it is contained within a windmill $W$;
(ii) we can analyze the memory regions within \textsf{VANE} using the
symbolic range analysis of Section~\ref{sub:sra};
(iii) \textsf{VANE} does not depend on another vane inside the same windmill $W$.

@@

The program in Figure~\ref{fig:ex_expansion} (a), which is a slight variation of
the code earlier seen in Figure~\ref{fig:ex_Regions}, contains two windmills.
The first consists in the \textsf{for} loop at line 1.
The second, nested in the first, is the \textsf{for} loop at line 2.
The vanes that sprout from these windmills, are highlighted in the program
dependence graph seen in Figure~\ref{fig:ex_expansion} (b).
The figure shows that we have three task candidates, each one formed by a
different vane.
In order to find the actual tasks, we apply function \textsf{expand}
(Fig.~\ref{fig:expand_alg}) onto the innermost vanes: \textsf{Vane-2} and
\textsf{Vane-3}.
The latter cannot be further expanded, because it depends on the former.
On the other hand, \textsf{Vane-2} can be expanded to include also
\textsf{Vane-3}, as this expansion does not create new dependences between vanes
in the same windmill.

\paragraph{Rational:}
We expand tasks up to a certain threshold \textsf{COST}, which is determined
by the cost model mentioned in Section~\ref{sub:profit}.
If the profitability of a task is less than this constant, then parallelism
might not pay off the cost of managing threads.
However, if we expand tasks too much, then we might not have much parallelism,
because code within a task is executed sequentially.
Thus, determining \textsf{COST} is essential to the performance of our
optimization.
In this work, we have determined this parameter empirically for the
runtime environment to be described in Section~\ref{sec:eval}.

\subsection{Privatization}
\label{sub:variance}

\Taskminer{} bestow a parallel semantics on code that has been originally
conceived to run sequentially.
This semantic gap might lead to race conditions.
Race conditions happen when annotations cause two or more threads to update
values that, in the sequential program, are denoted by the same name.
To avoid such situations, the insertion of annotations requires us to replicate
some values, making them private to each task.
We have named such replication {\em privatization}.

\begin{example}[Privatization]
\label{ex:priv}
Variables \textsf{i} and \textsf{j} in Figure~\ref{fig:ex_privatize} need to be
replicated among the tasks that the annotation at line 8 creates.
The need to replicate \textsf{j} is more apparent: this variable is updated in
the body of the task pragma, at line 9.
In the absence of replication, we have a race condition.
Variable \textsf{i} must also be replicated.
Each iteration of the loop at line 9 reads a different value of \textsf{i}.
Thus, each task should receive a different value in this variable.
In the absence of replication, all the threads would read the same value;  hence,
parallelization would change the semantics of the program.
\end{example}

\begin{definition}[The Private Requirement]
\label{def:private}
If a variable name has scalar type and contains at least one def-use chain that
enters the frontier of a task, then this name is said to bear the {\em private
requirement}.
The frontier of a task is determined by the \textsf{expand} function seen in
Figure~\ref{fig:expand_alg}.
\end{definition}

\begin{example}
\label{ex:priv_2}
Variable \textsf{i} must be privatized because it is defined at lines 2 and 10 of
Figure~\ref{fig:ex_privatize}, and is used at line 9 -- the body of a new task.
Moreover, this variable has a scalar type, e.g., \textsf{int}.
Similarly, variable \textsf{j} is defined at line 6, and is used at line 9.
This property -- the existence of def-use chains entering the task region --
leads to the insertion of the \textsf{firstprivate} clause at line 8 of
Figure~\ref{fig:ex_privatize}.
\end{example}

\noindent
\paragraph{firstprivate vs depend:} according to Definition~\ref{def:private},
only names representing scalar values are privatized.
These names have {\em value semantics}, i.e., they are ``passed
by copy" if used as actual arguments of functions.
Memory regions, that is to say, regions whose access happens through pointers,
are not privatized.
These regions are, instead, marked as dependences of the task, through the
\textsf{depend} clause.
Memory regions are discovered via the symbolic region analysis of
Section~\ref{sub:sra}.
In addition to pointers, we do not privatize values defined within
a task region, and used outside it.
That is to say: we privatize incoming def-use chains, but do not privatize
outgoing chains.
Values having this ``outgoing chain property" are shared by default.
We indicate this semantics via the \textsf{default(shared)} pragmas, as seen
in line 8 of Figure~\ref{fig:ex_privatize}.

@@

% Explain rational
The analyses described in Sections~\ref{sub:sra}-\ref{sub:variance} have
been implemented at the level of the LLVM intermediate representation.
We chose to implement those techniques at that level to benefit from the support
of several data-flow analyses already in place in the LLVM infra-structure.
In particular, this decision lets us reuse LLVM's scalar
evolution~\cite[p.18]{Grosser12}, from which we derive our symbolic analysis.
Additionally, we could reuse LLVM's program dependence graph, and its suite of
powerful alias analyses.
This toolbox gave us the necessary means to disambiguate pointers, albeit
conservatively, and estimate safe limits to memory regions.
However, our annotations are still inserted in C programs.
Having them in a high-level programming language has two main advantages:
they are human-readable, and thus can be debugged, and they are compatible with
compilers other than \textsf{clang}, such as \textsf{gcc}.
Thus, a major part of this work was the engineering effort to map
information from LLVM's intermediate representation back into C.

% Explain machinery: scope tree and simplification
\noindent
\textbf{The Scope Tree:}
To recover high-level information from the low-level IR, we have designed a
supporting data-structure henceforth named the {\em scope tree}.
The scope tree maps hammock regions into C constructions, such as
\textsf{while} and \textsf{if-then-else} blocks.
Each node of this graph represents a hammock region, augmented with
meta-information, such as the program part that it represents.
We keep track of these program parts via debugging information, which is
appended into the LLVM IR via the \textsf{-g} flag passed to the \textsf{clang}
compiler.
We have an edge from node $s_1$ to node $s_2$ if region $s_2$ is nested
within region $s_2$.
In the absence of control-flow optimizations such as loop unrolling or
dead-code elimination, each hammock region corresponds to some structured
code block (code region delimitable by braces).
Thus, we can find line numbers for each of the regions that we have marked
as tasks after the expansion seen in Section~\ref{sub:expansion}.

\noindent
\textbf{Simplification:}
before we annotate programs, we proceed to simplify these annotations.
Simplification happens via a system of rewriting rules, which explores 
identities.
Typical identifies include, for instance, $x + 0 = x$,
$\mathsf{max}(x, x) = x$ and $c \times \mathsf{max}(x, y) =
\mathsf{max}(c \times x, c \times y)$.
Identities are applied iteratively, until a fixed-point is reached.
Because we do not support commutativity, a fixed-point is guaranteed to be
always reached.
Notice that the source code that we produce is not totally equivalent to the
original program augmented with annotations.
To be able to insert annotations, we format the original code.
This operation involves, for instance, breaking lines containing multiple
statements, and inserting delimiting braces within every block in the program,
even one-liners.
This said, it is still possible that a task, after expansion, maps to a single
line that contains multiple statements, such as nested function calls.
In this case, we do not annotate the target program.
Section~\ref{sec:eval} provides data about \Taskminer's capacity to
annotate real-world programs.

\paragraph{Bounding Recursive Tasks}
Although the OpenMP runtime guarantees correctness, its scheduling policy
is heavily jeopardized by the excessive queueing up of tasks. Specifically, algorithms
such as \emph{SparseLU} and \emph{FFT} can suffer bloat because of
too many tasks in the Task Graph\cite{Huynh17}.

To solve this problem, we conservatively limit the number of tasks
that can be created tied to the same Task Region. We do that by two means:
(1) limiting the creation of recursive tasks by depth; (2) limiting the creation of 
tasks in the whole program. We do that by setting up static variables to count up
the depth level in recursive calls. When the depth reaches a certain level, which
can be defined by the user, task dispatch will be halted, avoiding creation of tasks that are
too small. This technique is illustrated in Figure \ref{fig:ex_cutoff}. 
This parameter's default value is very conservative and so can be easily switched off
or set by the user.
